---
title: "SMC Dataset Classification"
output: html_notebook
---

# General

Load necessary libraries.

```{r}
dir <- choose.dir() #Select file "R_final"
# setwd(dir)

# data manipulation
library(Seurat)
library(SeuratData) # devtools::install_github('satijalab/seurat-data')
library(data.table)
library(Matrix)
library(dplyr)
library(stringr)
library(readr)

# initial filtering
library(DropletUtils) #BiocManager::install("DropletUtils")

# gene set enrichment
#library(gprofiler2)

# plotting
library(ggplot2)
library(patchwork)
#library(gplots)
#library(cowplot)


set.seed(1)

#memory.limit(size = 70000)
```



# Step 1. Create a Seurat objects

The data was obtained from: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE132465
The data was not correctly formatted for Read10x function, so additional steps had to be performed once, following instructions here: https://www.biostars.org/p/449217/

```{r}
metadata <- fread(paste0(dir,"/Data/GSE132465_GEO_processed_CRC_10X_cell_annotation.txt.gz"))
metadata <- filter(metadata, metadata$Cell_type == "Epithelial cells")
rownames(metadata) <- metadata$Index
cols <- metadata$Index

# # Data has gene names as rows, and sample names/indexes as columns
# data <- fread(paste0(dir,"/Data/GSE132465_GEO_processed_CRC_10X_raw_UMI_count_matrix.txt.gz"), select = c("Index",cols))
# head(data[,1:10])
# sparse.mat <- Matrix(as.matrix(data[, -1, with=FALSE]), sparse=TRUE)
# DropletUtils::write10xCounts(paste0(dir, "/Data/GSE132465_SMC"), x = sparse.mat,
#                            gene.id = data$Index)
# rca.data <- DropletUtils::read10xCounts(paste0(dir, "/Data/GSE132465_SMC"), col.names=TRUE)

smc.data <- Read10X(data.dir = paste0(dir, "/Data/GSE132465_SMC"))
dim(smc.data)

smc_seurat <- CreateSeuratObject(smc.data, project = "smc_seurat", meta.data = metadata)
smc_seurat

smc_normal <- subset(smc_seurat, subset = Class == "Normal")
smc_tumour <- subset(smc_seurat, subset = Class == "Tumor")
dim(smc_normal)
dim(smc_tumour)

saveRDS(smc_normal, file = "./Results/SMC/smc_normal.RDS")
saveRDS(smc_tumour, file = "./Results/SMC/smc_tumour.RDS")
# smc_normal <- readRDS("./Results/SMC/smc_normal.RDS")
# smc_tumour <- readRDS("./Results/SMC/smc_tumour.RDS")
```



# Step 2. Check quality control


```{r, fig.height=10, fig.width=20}
smc_normal <- PercentageFeatureSet(smc_normal, col.name = "percent.mt", pattern = "^MT-")
smc_tumour <- PercentageFeatureSet(smc_tumour, col.name = "percent.mt", pattern = "^MT-")

VlnPlot(smc_normal, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
VlnPlot(smc_tumour, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```



# Step 3. Normalisation, feature selection, data scaling, PCA and UMAP

```{r}
smc_normal <- NormalizeData(smc_normal, normalization.method = "LogNormalize", scale.factor =   10000)
smc_normal <- FindVariableFeatures(smc_normal, selection.method = "vst", nfeatures = 3000)
smc_normal <- ScaleData(smc_normal, features = rownames(smc_normal))
smc_normal <- RunPCA(smc_normal, npcs = 50)
smc_normal <- RunUMAP(smc_normal, dims = 1:20)

smc_tumour <- NormalizeData(smc_tumour, normalization.method = "LogNormalize", scale.factor =   10000)
smc_tumour <- FindVariableFeatures(smc_tumour, selection.method = "vst", nfeatures = 3000)
smc_tumour <- ScaleData(smc_tumour, features = rownames(smc_tumour))
smc_tumour <- RunPCA(smc_tumour, npcs = 50)
smc_tumour <- RunUMAP(smc_tumour, dims = 1:20)

saveRDS(smc_normal, file = "./Results/SMC/smc_normal_2.RDS")
saveRDS(smc_tumour, file = "./Results/SMC/smc_tumour_2.RDS")
#smc_normal <- readRDS("./Results/SMC/smc_normal_2.RDS")
#smc_tumour <- readRDS("./Results/SMC/smc_tumour_2.RDS")
```



# Step 4. Classification

```{r}
library(scAnnotatR)

path_to_models = paste0(dir,"/scAnnotatR")
models <- load_models(path_to_models)
names(models)
```

```{r}
c_types <- c("Stem_TA cells","Intermediate cells","Colonocytes","BEST4 cells",
             "Goblet cells","Tuft cells")

seurat.obj.smc_normal <- classify_cells(classify_obj = smc_normal, 
                             assay = 'RNA', slot = 'data',
                             cell_types = c_types,
                             path_to_models = path_to_models)

seurat.obj.smc_tumour <- classify_cells(classify_obj = smc_tumour, 
                             assay = 'RNA', slot = 'data',
                             cell_types = c_types,
                             path_to_models = path_to_models)

saveRDS(seurat.obj.smc_normal, file = "./Results/SMC/seurat.obj.smc_normal.RDS")
saveRDS(seurat.obj.smc_tumour, file = "./Results/SMC/seurat.obj.smc_tumour.RDS")
#seurat.obj.smc_normal <- readRDS("./Results/SMC/seurat.obj.smc_normal.RDS")
#seurat.obj.smc_tumour <- readRDS("./Results/SMC/seurat.obj.smc_tumour.RDS")

smc_normal_class <- seurat.obj.smc_normal[[]][, c(11:ncol(seurat.obj.smc_normal[[]]))]
smc_normal_class

smc_tumour_class <- seurat.obj.smc_tumour[[]][, c(11:ncol(seurat.obj.smc_tumour[[]]))]
smc_tumour_class
```

```{r}
Seurat::DimPlot(seurat.obj.smc_normal, group.by = "most_probable_cell_type")
Seurat::DimPlot(seurat.obj.smc_tumour, group.by = "most_probable_cell_type")
```

```{r, fig.height=10,fig.width=20}
Seurat::DimPlot(seurat.obj.smc_normal, group.by = "predicted_cell_type")
Seurat::DimPlot(seurat.obj.smc_tumour, group.by = "predicted_cell_type")
```


```{r, fig.width=30, fig.height=20}
# Visualize the cell types
Seurat::DimPlot(seurat.obj.smc_normal, group.by = "predicted_cell_type",
                split.by = "predicted_cell_type", ncol = 5) + NoLegend()

Seurat::DimPlot(seurat.obj.smc_tumour, group.by = "predicted_cell_type",
                split.by = "predicted_cell_type", ncol = 5) + NoLegend()
```

```{r, fig.width=30, fig.height=30}
# Normal cells
# Merge hybrid and confused states
seurat.obj.smc_normal_alt <- seurat.obj.smc_normal
Idents(seurat.obj.smc_normal_alt) <- seurat.obj.smc_normal_alt@meta.data$predicted_cell_type
#levels(seurat.obj.smc_normal_alt)

new_ident <- setNames(c("Differentiating cells","Differentiating cells",
                        "Goblet cells","Stem-like/TA cells",
                        "Intermediate cells","Differentiating cells",
                        "Unknown","Differentiating cells",
                        "Differentiating cells","Differentiating cells",
                        "Differentiating cells","Colonocytes",
                        "Differentiating cells","Differentiating cells",
                        "BEST4+ cells","Differentiating cells",
                        "Confused cells","Differentiating cells",
                        "Hybrid cells","Tuft cells",
                        "Confused cells","Confused cells",
                        "Differentiating cells","Differentiating cells",
                        "Differentiating cells"),
                      levels(seurat.obj.smc_normal_alt))


seurat.obj.smc_normal_alt <- RenameIdents(seurat.obj.smc_normal_alt, new_ident)

my_levels <- c("Stem-like/TA cells","Intermediate cells",
               "Colonocytes","BEST4+ cells","Goblet cells","Tuft cells","Differentiating cells",
               "Hybrid cells","Confused cells","Unknown")

# Relevel object@ident
Idents(seurat.obj.smc_normal_alt) <- factor(Idents(seurat.obj.smc_normal_alt), levels = my_levels)
seurat.obj.smc_normal_alt$orig.ident <- Idents(seurat.obj.smc_normal_alt)



# Tumour cells
# Merge hybrid and confused states
seurat.obj.smc_tumour_alt <- seurat.obj.smc_tumour
Idents(seurat.obj.smc_tumour_alt) <- seurat.obj.smc_tumour_alt@meta.data$predicted_cell_type
#levels(seurat.obj.smc_tumour_alt)

new_ident <- setNames(c("Stem-like/TA cells","Unknown",
                        "Goblet cells","Intermediate cells",
                        "Differentiating cells","Differentiating cells",
                        "Differentiating cells","Differentiating cells",
                        "Differentiating cells","BEST4+ cells",
                        "Differentiating cells","Differentiating cells",
                        "Colonocytes","Confused cells",
                        "Confused cells","Differentiating cells",
                        "Differentiating cells","Hybrid cells",
                        "Differentiating cells","Differentiating cells",
                        "Differentiating cells","Differentiating cells",
                        "Differentiating cells","Confused cells",
                        "Differentiating cells","Confused cells",
                        "Differentiating cells","Differentiating cells",
                        "Differentiating cells","Differentiating cells",
                        "Differentiating cells","Differentiating cells",
                        "Differentiating cells","Confused cells",
                        "Differentiating cells","Confused cells",
                        "Differentiating cells","Tuft cells",
                        "Confused cells","Differentiating cells",
                        "Differentiating cells","Differentiating cells",
                        "Hybrid cells","Differentiating cells",
                        "Differentiating cells"),
                      levels(seurat.obj.smc_tumour_alt))


seurat.obj.smc_tumour_alt <- RenameIdents(seurat.obj.smc_tumour_alt, new_ident)

my_levels <- c("Stem-like/TA cells","Intermediate cells",
               "Colonocytes","BEST4+ cells","Goblet cells","Tuft cells","Differentiating cells",
               "Hybrid cells","Confused cells","Unknown")

# Relevel object@ident
Idents(seurat.obj.smc_tumour_alt) <- factor(Idents(seurat.obj.smc_tumour_alt), levels = my_levels)
seurat.obj.smc_tumour_alt$orig.ident <- Idents(seurat.obj.smc_tumour_alt)
```

```{r, fig.height=5, fig.width=8}
Seurat::DimPlot(seurat.obj.smc_normal_alt)
Seurat::DimPlot(seurat.obj.smc_normal_alt, split.by = "orig.ident", ncol = 5) + NoLegend()

Seurat::DimPlot(seurat.obj.smc_tumour_alt)
Seurat::DimPlot(seurat.obj.smc_tumour_alt, split.by = "orig.ident", ncol = 5) + NoLegend()
```

```{r, fig.height=20, fig.width=20}
markers_int <- read.table(file="./Results/Normal Cells/markers_int.txt")
markers.to.plot <- markers_int[,1]
#markers.to.plot <- unique(unlist(strsplit(markers.to.plot, " ")))

DoHeatmap(seurat.obj.smc_normal_alt, features = markers.to.plot, assay = "RNA")
DoHeatmap(seurat.obj.smc_tumour_alt, features = markers.to.plot, assay = "RNA")
```

```{r, fig.height=8, fig.width=20}
markers.to.plot <- unique(unlist(strsplit(markers.to.plot, " ")))

DotPlot(seurat.obj.smc_normal_alt, features = markers.to.plot,
        #group.by = "predicted_cell_type",
        dot.scale = 8, cols = c("blue","red")) + 
  theme(axis.text.x = element_text(angle = 45, hjust=1))

DotPlot(seurat.obj.smc_tumour_alt, features = markers.to.plot,
        #group.by = "predicted_cell_type",
        dot.scale = 8, cols = c("blue","red")) + 
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```


```{r}
df_class <- data.frame(Class = c(seurat.obj.smc_normal_alt$Class,seurat.obj.smc_tumour_alt$Class),
                       Cell_Types = c(seurat.obj.smc_normal_alt$orig.ident, 
                                      seurat.obj.smc_tumour_alt$orig.ident))

df_class[,1] <- ifelse(df_class[,1] == "Tumor", "Tumour", df_class[,1])

total_class <- plyr::count(df_class, vars = "Class")
prct_class <- plyr::count(df_class, vars = c("Class","Cell_Types"))
prct_class <- merge(prct_class,total_class, by = "Class")
prct_class$Percentage <- (prct_class$freq.x / prct_class$freq.y) * 100

ggplot(prct_class, aes(fill = Cell_Types, y = Percentage, x = Class)) + 
    geom_bar(position="fill", stat="identity") 
```

```{r,fig.height=5,fig.width=5}
df_class_normal <- data.frame(Class = c(seurat.obj.smc_normal_alt$Class),
                       Cell_Types = c(seurat.obj.smc_normal_alt$orig.ident))

df_class_normal[,1] <- ifelse(df_class_normal[,1] == "Tumor", "Tumour", df_class_normal[,1])

total_class <- plyr::count(df_class_normal, vars = "Class")
prct_class <- plyr::count(df_class_normal, vars = c("Class","Cell_Types"))
prct_class <- merge(prct_class,total_class, by = "Class")
prct_class$Percentage <- (prct_class$freq.x / prct_class$freq.y) * 100

ggplot(prct_class, aes(fill = Cell_Types, y = Percentage, x = Class)) + 
    geom_bar(position="fill", stat="identity") 
```

```{r,fig.height=5,fig.width=5}
df_class_tumour <- data.frame(Class = c(seurat.obj.smc_tumour_alt$Class),
                       Cell_Types = c(seurat.obj.smc_tumour_alt$orig.ident))

df_class_tumour[,1] <- ifelse(df_class_tumour[,1] == "Tumor", "Tumour", df_class_tumour[,1])

total_class <- plyr::count(df_class_tumour, vars = "Class")
prct_class <- plyr::count(df_class_tumour, vars = c("Class","Cell_Types"))
prct_class <- merge(prct_class,total_class, by = "Class")
prct_class$Percentage <- (prct_class$freq.x / prct_class$freq.y) * 100

ggplot(prct_class, aes(fill = Cell_Types, y = Percentage, x = Class)) + 
    geom_bar(position="fill", stat="identity") 
```

```{r}
df_conf <- data.frame(Class = c(seurat.obj.smc_normal_alt$Class,seurat.obj.smc_tumour_alt$Class),
                      Cell_Types = c(seurat.obj.smc_normal_alt$orig.ident,
                                      seurat.obj.smc_tumour_alt$orig.ident),
                      scAnnotatR_Types = c(seurat.obj.smc_normal_alt$predicted_cell_type,
                                      seurat.obj.smc_tumour_alt$predicted_cell_type))
df_conf <- filter(df_conf, Cell_Types == "Confused cells")

total_conf <- plyr::count(df_conf, vars = "Class")
prct_conf <- plyr::count(df_conf, vars = c("Class","scAnnotatR_Types"))
prct_conf <- merge(prct_conf,total_conf, by = "Class")
prct_conf$Percentage <- (prct_conf$freq.x / prct_conf$freq.y) * 100

ggplot(prct_conf, aes(fill = scAnnotatR_Types, y = Percentage, x = Class)) +
    geom_bar(position="fill", stat="identity")
```

We want to calculate the percentage of each cell type per CMS.

```{r}
df_cms <- data.frame(CMS = seurat.obj.smc_tumour_alt$Cell_subtype,
                     Cell_Types = seurat.obj.smc_tumour_alt$orig.ident)

total_cms <- plyr::count(df_cms, vars = "CMS")
prct_cms <- plyr::count(df_cms, vars = c("CMS","Cell_Types"))
prct_cms <- merge(prct_cms,total_cms, by = "CMS")
prct_cms$Percentage <- (prct_cms$freq.x / prct_cms$freq.y) * 100

ggplot(prct_cms, aes(fill = Cell_Types, y = Percentage, x = CMS)) + 
    geom_bar(position="fill", stat="identity") 
```

Check in some of the unknown cells are EECs.

```{r, fig.height=8, fig.width=10}
eec_markers <- c("CHGA","CHGB","NEUROD1","NEUROG3")

unknown_normal <- subset(seurat.obj.smc_normal_alt, ident = "Unknown")
unknown_tumour <- subset(seurat.obj.smc_tumour_alt, ident = "Unknown")
  
FeaturePlot(unknown_normal, features = eec_markers, ncol = 2)
FeaturePlot(unknown_tumour, features = eec_markers, ncol = 2)
```


