---
title: "Classification of Tumour Epithelial Cells"
author: "Francisca Xara-Brasil"
output: html_notebook
---

Tutorial: https://github.com/grisslab/scAnnotatR/blob/master/vignettes/classifying-cells.Rmd


# Installation

The `scAnnotatR` package can be directly installed from Bioconductor:

```{r, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if (!require(scAnnotatR))
  BiocManager::install("scAnnotatR")
```

For more information, see https://bioconductor.org/install/.

```{r}
# load scAnnotatR into working space
library(scAnnotatR)
library(dplyr)
library(writexl)
library(ggplot2)

dir <- choose.dir() #Select file "R_final"
```



# Models

```{r}
path_to_models = paste0(dir,"/scAnnotatR")
models <- load_models(path_to_models)
names(models)
```

```{r}
models[['Stem_TA cells']]
models[['Intermediate cells']]
models[['Colonocytes']]
models[['BEST4 cells']]
models[['Goblet cells']]
models[['Tuft cells']]
```



# Basic pipeline to identify cell types in a scRNA-seq dataset using scAnnotatR   

## Preparing the data

```{r}
# load the example dataset
ep_tumour <- readRDS(paste0(dir,"/Results/Tumour Cells/ep_tumour_2.RDS"))
ep_tumour
```

The example dataset already contains the clustering results as part of the metadata. This is **not** necessary for the classification process.

```{r}
head(ep_tumour[[]])
```



## Cell classification

To launch cell type identification, we simply call the `classify_cells` 
function. A detailed description of all parameters can be found through
the function's help page `?classify_cells`.

To use all pretrained classifiers on the dataset, `cell_types = 'all'` can be used.

```{r}
c_types <- c("Stem_TA cells","Intermediate cells","Colonocytes","BEST4 cells",
             "Goblet cells","Tuft cells")

tumour_class.obj <- classify_cells(classify_obj = ep_tumour, 
                             assay = 'RNA', slot = 'data',
                             cell_types = c_types,
                             path_to_models = path_to_models)

saveRDS(tumour_class.obj, file = paste0(dir,"/scAnnotatR/tumour_class.obj.RDS"))
#tumour_class.obj <- readRDS(paste0(dir,"/scAnnotatR/tumour_class.obj.RDS"))
```



## Parameters

  * The option **cell_types = 'all'** tells the function to use all available
    cell classification models.
    Alternatively, we can limit the identifiable cell types: 
    * by specifying: `cell_types = c('Stem_TA cells', 'Intermediate cells')`
    * or by indicating the applicable classifier using the **classifiers** option: 
      `classifiers = c(default_models[['Stem_TA cells']], default_models[['Intermediate cells']])`

  * The option **path_to_models = 'default'** is to automatically use the 
    package-integrated pretrained models (without loading the models into the
    current working space). This option can be used to load a local database instead.
    For more details see the vignettes on training your own classifiers.



# Result interpretation

The `classify_cells` function returns the input object
but with additional columns in the metadata table.

```{r}
# display the additional metadata fields
tumour_class <- tumour_class.obj[[]][, c(22:ncol(tumour_class.obj[[]]))]
tumour_class

# write.table(tumour_class,
#             file=paste0(dir,"/scAnnotatR/scAnnotatR_tumour_class.txt"),
#             row.names=TRUE, col.names=TRUE)
```

New columns are:

  * **predicted_cell_type**: The predicted cell type, also containing any 
    ambiguous assignments. In these cases, the possible cell types are separated
    by a "/"

  * **most_probable_cell_type**: contains the most probably cell type ignoring any 
    ambiguous assignments.

  * columns with syntax `[celltype]_p`: probability of a cell to belong 
    to a cell type. Unknown cell types are marked as NAs.



## Result visualization

The predicted cell types can now simply be visualized using the matching
plotting functions. 

```{r}
# Visualize the cell types
Seurat::DimPlot(tumour_class.obj, group.by = "most_probable_cell_type")
Seurat::DimPlot(tumour_class.obj, group.by = "most_probable_cell_type", split.by = "Patient",
                ncol = 3)
Seurat::DimPlot(tumour_class.obj, group.by = "most_probable_cell_type", split.by = "Cell_subtype", 
                ncol = 2)
```

```{r, fig.width=30, fig.height=10}
# Visualize the cell types
Seurat::DimPlot(tumour_class.obj, group.by = "predicted_cell_type")
```

```{r, fig.width=20, fig.height=25}
# Visualize the cell types
Seurat::DimPlot(tumour_class.obj, group.by = "predicted_cell_type",
                split.by = "predicted_cell_type", ncol = 4) + NoLegend()
```

```{r, fig.width=30, fig.height=30}
# Merge hybrid and confused states
tumour_class.obj_alt <- tumour_class.obj
Idents(tumour_class.obj_alt) <- tumour_class.obj_alt@meta.data$predicted_cell_type
#levels(tumour_class.obj_alt)

new_ident <- setNames(c("Goblet cells", "Stem-like/TA cells",
                        "Differentiating cells","Unknown",
                        "BEST4+ cells","Colonocytes",
                        "Intermediate cells","Confused cells",
                        "Differentiating cells","Differentiating cells",
                        "Differentiating cells","Differentiating cells",
                        "Differentiating cells","Confused cells",
                        "Differentiating cells","Differentiating cells",
                        "Differentiating cells","Differentiating cells",
                        "Differentiating cells","Differentiating cells",
                        "Differentiating cells","Differentiating cells",
                        "Tuft cells","Differentiating cells",
                        "Differentiating cells","Differentiating cells",
                        "Hybrid cells","Differentiating cells",
                        "Differentiating cells","Hybrid cells",
                        "Differentiating cells","Differentiating cells",
                        "Confused cells","Confused cells",
                        "Differentiating cells","Differentiating cells"),
                      levels(tumour_class.obj_alt))

tumour_class.obj_alt <- RenameIdents(tumour_class.obj_alt, new_ident)

my_levels <- c("Stem-like/TA cells","Intermediate cells",
               "Colonocytes","BEST4+ cells","Goblet cells","Tuft cells",
               "Differentiating cells","Hybrid cells",
               "Confused cells","Unknown")

# Relevel object@ident
Idents(tumour_class.obj_alt) <- factor(Idents(tumour_class.obj_alt), levels = my_levels)
tumour_class.obj_alt$orig.ident <- Idents(tumour_class.obj_alt)
```

```{r, fig.height=4, fig.width=7}
Seurat::DimPlot(tumour_class.obj_alt)
Seurat::DimPlot(tumour_class.obj_alt, split.by = "orig.ident", ncol = 5) + NoLegend()
```


```{r, fig.height=20, fig.width=23}
markers_int <- read.table(file=paste0(dir,"/Results/Normal Cells/markers_int.txt"))
markers.to.plot <- markers_int[,1]

DoHeatmap(tumour_class.obj_alt, features = markers.to.plot, assay = "RNA",
          group.by = "most_probable_cell_type") 
DoHeatmap(tumour_class.obj_alt, features = markers.to.plot, assay = "RNA")
```

```{r, fig.height=8, fig.width=20}
markers.to.plot <- unique(unlist(strsplit(markers.to.plot, " ")))

DotPlot(tumour_class.obj_alt, features = markers.to.plot, assay = "RNA",
        #group.by = "predicted_cell_type",
        dot.scale = 8, cols = c("blue","red")) + 
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```

For a certain cell type, users can also view the prediction probability.

```{r fig.width = 10, fig.height = 10}
# Visualize the cell types
Seurat::FeaturePlot(tumour_class.obj_alt,
                    features = c("Stem_TA_cells_p","Intermediate_cells_p",
                                 "Colonocytes_p","Goblet_cells_p","Tuft_cells_p"))
```

We want to calculate the percentage of each cell type in the whole sample.

```{r, fig.height=5,fig.width=5}
df_class <- data.frame(Class = tumour_class.obj_alt$Class,
                       Cell_Types = tumour_class.obj_alt$orig.ident)

df_class[,1] <- ifelse(df_class[,1] == "Tumor", "Tumour", df_class[,1])

total_class <- plyr::count(df_class, vars = "Class")
prct_class <- plyr::count(df_class, vars = c("Class","Cell_Types"))
prct_class <- merge(prct_class,total_class, by = "Class")
prct_class$Percentage <- (prct_class$freq.x / prct_class$freq.y) * 100

ggplot(prct_class, aes(fill = Cell_Types, y = Percentage, x = Class)) + 
    geom_bar(position="fill", stat="identity")
```

Per patient.

```{r}
df_patient <- data.frame(Patients = tumour_class.obj_alt$Patient,
                         Cell_Types = tumour_class.obj_alt$orig.ident)

total_patient <- plyr::count(df_patient, vars = "Patients")
prct_patient <- plyr::count(df_patient, vars = c("Patients","Cell_Types"))
prct_patient <- merge(prct_patient,total_patient, by = "Patients")
prct_patient$Percentage <- (prct_patient$freq.x / prct_patient$freq.y) * 100


ggplot(prct_patient, aes(fill = Cell_Types, y = Percentage, x = Patients)) + 
    geom_bar(position="fill", stat="identity") 
```

We want to calculate the percentage of each cell type per CMS.

```{r}
df_cms <- data.frame(CMS = tumour_class.obj_alt$Cell_subtype,
                         Cell_Types = tumour_class.obj_alt$orig.ident)

total_cms <- plyr::count(df_cms, vars = "CMS")
prct_cms <- plyr::count(df_cms, vars = c("CMS","Cell_Types"))
prct_cms <- merge(prct_cms,total_cms, by = "CMS")
prct_cms$Percentage <- (prct_cms$freq.x / prct_cms$freq.y) * 100

ggplot(prct_cms, aes(fill = Cell_Types, y = Percentage, x = CMS)) + 
    geom_bar(position="fill", stat="identity") 
```

Plot a Sankey diagram.

```{r}
library(networkD3)
library(tidyr)
library(tibble)

# Create an incidence matrix. Usually the flow goes from the row names to the column names.
# Remember that our connection are directed since we are working with a flow.
mtx <- as.matrix(dcast(prct_cms, CMS ~ Cell_Types, value.var="freq.x", fill=0))
rownames(mtx) <- mtx[,1]
mtx <- mtx[,-1]

# Transform it to connection data frame with tidyr from the tidyverse:
links <- mtx %>% 
  as.data.frame() %>% 
  rownames_to_column(var="source") %>% 
  gather(key="target", value="value", -1) %>%
  filter(value != 0)
 
# From these flows we need to create a node data frame: it lists every entities involved in the flow
nodes <- data.frame(
  name=c(as.character(links$source), as.character(links$target)) %>% 
    unique()
  )
 
# With networkD3, connection must be provided using id, not using real name like in the links dataframe.. So we need to reformat it.
links$IDsource <- match(links$source, nodes$name)-1 
links$IDtarget <- match(links$target, nodes$name)-1

# prepare colour scale
ColourScal ='d3.scaleOrdinal() .range([""])'

my_color <- 'd3.scaleOrdinal() .domain(["CMS1", "CMS2", "CMS3", "CMS4", "Stem-like/TA", "Intermediate", "Colonocytes", "BEST4+", "Goblet", "Tuft", "Differentiating", "Hybrid", "Confused", "Unknown"]) .range(["#F8766D", "#E68613", "#ABA300", "#0CB702", "#F8766D", "#E68613", "#ABA300", "#0CB702", "#00BE67", "#00C19A", "#00A9FF", "#8494FF", "#ED68ED", "#FF68A1"])'
 
 
# Make the Network
p <- sankeyNetwork(Links = links, Nodes = nodes,
                     Source = "IDsource", Target = "IDtarget",
                     Value = "value", NodeID = "name", 
                     sinksRight=FALSE, colourScale = my_color,fontSize = 12)

p
```



## Conectivity Analysis with PAGA

To evaluate how different cell types are connected with each other and to understand potential developmental relationships, we will use PAGA, which is included as part of the scanpy toolkit in Python.

As scanpy is a python package and does not support a Seurat object as the input, we need to store its information in a format that scanpy supports. We will use the anndata package to create a h5ad file with the information needed (e.g. PCA, UMAP, and cell type annotation).

```{r}
library(anndata) #install.packages("anndata")
adata <- AnnData(X = t(as.matrix(tumour_class.obj_alt[['RNA']]@data)),
                 obs = data.frame(celltype = tumour_class.obj_alt@active.ident,
                                  row.names = colnames(tumour_class.obj_alt)),
                 var = tumour_class.obj_alt[['RNA']]@meta.features,
                 layers = list(counts = t(as.matrix(tumour_class.obj_alt[['RNA']]@counts))),
                 obsm = list(pca = Embeddings(tumour_class.obj_alt,"pca")[,1:20],
                             umap = Embeddings(tumour_class.obj_alt,"umap"))
                )
adata$write_h5ad(paste0(dir,"./Results/Tumour Cells/ep_tumour_anndata_obj.h5ad"))
```

Next, we need to swtich to Python (>=3.6). To stay in R, we will use the reticulate R package, which provides an interface to Python.

```{r}
library(reticulate)
py_install("scanpy", pip=T)
sc <- import("scanpy")
```

PAGA analysis, using the results previously obtained with Seurat.

```{r}
adata_ep_tumour <- sc$read(paste0(dir,"./Results/Tumour Cells/ep_tumour_anndata_obj.h5ad")) # load the h5ad file

sc$pp$neighbors(adata_ep_tumour, n_neighbors=20L, use_rep='pca')
sc$tl$paga(adata_ep_tumour, groups='celltype')
adata_ep_tumour$write_h5ad(paste0(dir,"./Results/Tumour Cells/ep_tumour_anndata_obj.h5ad"))
```

Plot the PAGA estimated cluster connectivity.

```{r}
plt <- import("matplotlib")
plt$use("Agg", force = TRUE)
sc$pl$paga(adata_ep_tumour,
           color='celltype',
           fontsize=7,
           fontoutline = 2,
           node_size_scale = 1,
           edge_width_scale = 0.5,
           frameon=FALSE,
           save="ep_tumour.png")
```


