---
title: "Training New Classification Models"
author: "Francisca Xara-Brasil"
output: html_notebook
---

Tutorial: https://github.com/grisslab/scAnnotatR/blob/master/vignettes/training-basic-model.Rmd


# Preparing train object and test object

To start the training workflow, we first install and load the necessary libraries. 

```{r, eval = FALSE}
# use BiocManager to install from Bioconductor
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
# the scAnnotatR package
if (!require(scAnnotatR))
  BiocManager::install("scAnnotatR")
# we use the scRNAseq package to load example data
# if (!require(scRNAseq))
#   BiocManager::install("scRNAseq")
```

```{r}
dir <- choose.dir() #Select file "R_final"
setwd(dir)

library(scAnnotatR)
```

First, we load the dataset that will be used for training (data from normal epithelial cells).

```{r}
ep_normal <- readRDS(file = paste0(dir,"/scAnnotatR/ep_normal_final.RDS"))
```

We split this dataset into two parts, one for the training and the other for the testing.

```{r}
pivot <- ncol(ep_normal)%/%2
train_set <- ep_normal[, 1:pivot]
test_set <- ep_normal[, (1+pivot):ncol(ep_normal)]
```


We want to train a classifier for each epithelial cell type (6 classifiers in total).

Considering, for instance, the Stem-like/TA cell classifier, we label all the cells that are not from that type as "others".

```{r}
# change cell label
train_set$Stem <- unlist(lapply(Idents(ep_normal),
                                  function(x) if (is.na(x)) {'ambiguous'} else if (x %in% c('Stem_TA cells')) {'Stem_TA cells'} else {'others'}))
test_set$Stem <- unlist(lapply(Idents(ep_normal),
                                 function(x) if (is.na(x)) {'ambiguous'} else if (x %in% c('Stem_TA cells')) {'Stem_TA cells'} else {'others'}))


train_set$Intermediate <- unlist(lapply(Idents(ep_normal),
                                  function(x) if (is.na(x)) {'ambiguous'} else if (x %in% c('Intermediate cells')) {'Intermediate cells'} else {'others'}))
test_set$Intermediate <- unlist(lapply(Idents(ep_normal),
                                 function(x) if (is.na(x)) {'ambiguous'} else if (x %in% c('Intermediate cells')) {'Intermediate cells'} else {'others'}))


train_set$Colonocyte <- unlist(lapply(Idents(ep_normal),
                                  function(x) if (is.na(x)) {'ambiguous'} else if (x %in% c('Colonocytes')) {'Colonocytes'} else {'others'}))
test_set$Colonocyte <- unlist(lapply(Idents(ep_normal),
                                 function(x) if (is.na(x)) {'ambiguous'} else if (x %in% c('Colonocytes')) {'Colonocytes'} else {'others'}))


train_set$BEST4 <- unlist(lapply(Idents(ep_normal),
                                  function(x) if (is.na(x)) {'ambiguous'} else if (x %in% c('BEST4 cells')) {'BEST4 cells'} else {'others'}))
test_set$BEST4 <- unlist(lapply(Idents(ep_normal),
                                 function(x) if (is.na(x)) {'ambiguous'} else if (x %in% c('BEST4 cells')) {'BEST4 cells'} else {'others'}))


train_set$Goblet <- unlist(lapply(Idents(ep_normal),
                                  function(x) if (is.na(x)) {'ambiguous'} else if (x %in% c('Goblet cells')) {'Goblet cells'} else {'others'}))
test_set$Goblet <- unlist(lapply(Idents(ep_normal),
                                 function(x) if (is.na(x)) {'ambiguous'} else if (x %in% c('Goblet cells')) {'Goblet cells'} else {'others'}))


train_set$Tuft <- unlist(lapply(Idents(ep_normal),
                                  function(x) if (is.na(x)) {'ambiguous'} else if (x %in% c('Tuft cells')) {'Tuft cells'} else {'others'}))
test_set$Tuft <- unlist(lapply(Idents(ep_normal),
                                 function(x) if (is.na(x)) {'ambiguous'} else if (x %in% c('Tuft cells')) {'Tuft cells'} else {'others'}))
```

We may want to check the number of cells in each category:

```{r}
table(train_set$Stem)
table(test_set$Stem)
table(train_set$Intermediate)
table(test_set$Intermediate)
table(train_set$Colonocyte)
table(test_set$Colonocyte)
table(train_set$BEST4)
table(test_set$BEST4)
table(train_set$Goblet)
table(test_set$Goblet)
table(train_set$Tuft)
table(test_set$Tuft)
```



# Setting marker genes

Next, we set the marker genes of each cell type (previously determined), which will be used for training the respective classification model. 

```{r}
markers_int <- read.table(file=paste0(dir,"/Results/Normal Cells/markers_int.txt"))

Stem_markers <- dplyr::filter(markers_int, Cell_Types == "Stem-like/TA cells")[,1]
Intermediate_markers <- dplyr::filter(markers_int, Cell_Types == "Intermediate cells")[,1]
Colonocytes_markers <- dplyr::filter(markers_int, Cell_Types == "Colonocytes")[,1]
BEST4_markers <- dplyr::filter(markers_int, Cell_Types == "BEST4+ cells")[,1]
Goblet_markers <- dplyr::filter(markers_int, Cell_Types == "Goblet cells")[,1]
Tuft_markers <- dplyr::filter(markers_int, Cell_Types == "Tuft cells")[,1]
```



# Train model

When the model is being trained, three pieces of information must be 
provided: 

  * the Seurat object used for training
  * the set of applied marker genes
  * the cell type defining the trained model

In case the dataset does not contain any cell classified as the target
cell type, the function will fail.  

When training on an imbalanced dataset (f.e. a dataset containing 90% Stem cells and
only very few other cell types), the trained model may bias toward the 
majority group and ignore the presence of the minority group. To avoid this, 
the number of positive cells and negative cells will be automatically balanced 
before training. Therefore, a smaller number cells will be randomly picked  
from the majority group. To use the same set of cells while training multiple 
times for one model, users can use `set.seed`. 

```{r}
set.seed(123)
classifier_Stem <- train_classifier(train_obj = train_set, cell_type = "Stem_TA cells", 
                                 marker_genes = Stem_markers,
                                 assay = 'RNA', slot = "data",
                                 tag_slot = 'active.ident')

classifier_Intermediate <- train_classifier(train_obj = train_set,
                                 cell_type = "Intermediate cells", 
                                 marker_genes = Intermediate_markers,
                                 assay = 'RNA', slot = "data",
                                 tag_slot = 'active.ident')

classifier_Colonocyte <- train_classifier(train_obj = train_set,
                                 cell_type = "Colonocytes", 
                                 marker_genes = Colonocytes_markers,
                                 assay = 'RNA', slot = "data",
                                 tag_slot = 'active.ident')

classifier_BEST4 <- train_classifier(train_obj = train_set,
                                 cell_type = "BEST4 cells", 
                                 marker_genes = BEST4_markers,
                                 assay = 'RNA', slot = "data",
                                 tag_slot = 'active.ident')

classifier_Goblet <- train_classifier(train_obj = train_set, cell_type = "Goblet cells", 
                                 marker_genes = Goblet_markers,
                                 assay = 'RNA', slot = "data",
                                 tag_slot = 'active.ident')

classifier_Tuft <- train_classifier(train_obj = train_set, cell_type = "Tuft cells", 
                                 marker_genes = Tuft_markers,
                                 assay = 'RNA', slot = "data",
                                 tag_slot = 'active.ident')
```


```{r}
classifier_Stem
classifier_Intermediate
classifier_Colonocyte
classifier_BEST4
classifier_Goblet
classifier_Tuft
```

The classification model is a `scAnnotatR` object. 
Details about the classification model are accessible via getter methods. 

For example:

```{r}
caret_model(classifier_Stem)
caret_model(classifier_Intermediate)
caret_model(classifier_Colonocyte)
caret_model(classifier_BEST4)
caret_model(classifier_Goblet)
caret_model(classifier_Tuft)
```



# Test model

The `test_classifier` model automatically tests a classifier's performance
against another dataset. Here, we used the `test_set` created before:

```{r}
print("Stem/TA")
classifier_Stem_test <- test_classifier(classifier = classifier_Stem,
                                        test_obj = test_set,
                                        assay = 'RNA', slot = "data",
                                        tag_slot = 'active.ident')

print("Intermediate")
classifier_Intermediate_test <- test_classifier(classifier = classifier_Intermediate,
                                        test_obj = test_set,
                                        assay = 'RNA', slot = "data",
                                        tag_slot = 'active.ident')

print("Colonocyte")
classifier_Colonocyte_test <- test_classifier(classifier = classifier_Colonocyte,
                                        test_obj = test_set,
                                        assay = 'RNA', slot = "data",
                                        tag_slot = 'active.ident')

print("BEST4")
classifier_BEST4_test <- test_classifier(classifier = classifier_BEST4,
                                        test_obj = test_set,
                                        assay = 'RNA', slot = "data",
                                        tag_slot = 'active.ident')
                                  
print("Goblet")
classifier_Goblet_test <- test_classifier(classifier = classifier_Goblet,
                                        test_obj = test_set,
                                        assay = 'RNA', slot = "data",
                                        tag_slot = 'active.ident')

print("Tuft")
classifier_Tuft_test <- test_classifier(classifier = classifier_Tuft,
                                        test_obj = test_set,
                                        assay = 'RNA', slot = "data",
                                        tag_slot = 'active.ident')

```


## Interpreting test model result

Apart from the output exported to console, test classifier function also returns an object, which is a list of:
  
  * **test_tag**: actual cell label, this can be different from the label provided by users because of ambiguous characters or the incoherence in cell type and sub cell type label assignment.  

  * **pred**: cell type prediction using current classifier

  * **acc**: prediction accuracy at the fixed probability threshold, the probability threshold value can also be queried using *p_thres(classifier)*

  * **auc**: AUC score provided by current classifier
  
  * **overall_roc**: True Positive Rate and False Positive Rate with a certain number of prediction probability thresholds
  
Every classifier internally consists of a trained SVM and a probability threshold. Only cells that are classified with a probability exceeding this threshold are classified as the respective cell type. The *overall_roc* slot summarizes the True Positive Rate (sensitivity) and False Positive Rate (1 - specificity) obtained by the trained model according to different thresholds.

```{r}
print("Stem/TA")
classifier_Stem_test$overall_roc
classifier_Stem_test$auc

print("Intermediate")
classifier_Intermediate_test$overall_roc
classifier_Intermediate_test$auc

print("Colonocytes")
classifier_Colonocyte_test$overall_roc
classifier_Colonocyte_test$auc

print("BEST4")
classifier_BEST4_test$overall_roc
classifier_BEST4_test$auc

print("Goblet")
classifier_Goblet_test$overall_roc
classifier_Goblet_test$auc

print("Tuft")
classifier_Tuft_test$overall_roc
classifier_Tuft_test$auc
```

The default probability threshold is 0.5. The higher sensitivity can be reached if we set the p_thres at 0.4. However, we will then have lower specificity, which means that we will incorrectly classify some cells as, for example, stem cells. At the same time, we may not retrieve all actual stem cells with higher p_thres (0.6, for example).

There is of course a certain trade-off between the sensitivity and the specificity of the model. Depending on the need of the project or the user-own preference, a probability threshold giving higher sensitivity or higher specificity can be chosen.

Since we want to guarantee that a low number of cells are misclassified rather than detecting all positive cells (we are more interested in high specificity rather than high sensitivity), we select, for each model, the lowest probability threshold with an associated FPR below 0.1.

```{r}
#p_thres(classifier_Stem) <- 0.6
p_thres(classifier_Intermediate) <- 0.4
p_thres(classifier_Colonocyte) <- 0.7
p_thres(classifier_BEST4) <- 0.4
#p_thres(classifier_Goblet) <- 0.3
#p_thres(classifier_Tuft) <- 0.5
```



## Plotting ROC curve

Apart from numbers, scAnnotatR also provide a method to plot the ROC curve. 

```{r}
roc_curve_Stem <- plot_roc_curve(test_result = classifier_Stem_test) +
             labs(title = "Stem-like/TA cells")

roc_curve_Intermediate <- plot_roc_curve(test_result = classifier_Intermediate_test) +
             labs(title = "Intermediate cells")

roc_curve_Colonocyte <- plot_roc_curve(test_result = classifier_Colonocyte_test) +
             labs(title = "Colonocytes")

roc_curve_BEST4 <- plot_roc_curve(test_result = classifier_BEST4_test) +
             labs(title = "BEST4+ cells")

roc_curve_Goblet <- plot_roc_curve(test_result = classifier_Goblet_test) +
             labs(title = "Goblet cells")

roc_curve_Tuft <- plot_roc_curve(test_result = classifier_Tuft_test) +
             labs(title = "Tuft cells")
```


```{r, fig.height=6, fig.width=10}
(roc_curve_Stem + roc_curve_Intermediate + roc_curve_Colonocyte) /
(roc_curve_BEST4 + roc_curve_Goblet + roc_curve_Tuft)
```



## Which model to choose?

Changes in the training data, in the set of marker genes and in the prediction 
probability threshold will all lead to a change in model performance.

There are several ways to evaluate the trained model, including the overall 
accuracy, the AUC score and the sensitivity/specificity of the model when 
testing on an independent dataset.

*Tip: Using more general markers of the whole population leads to higher 
sensitivity. This sometimes produces lower specificity because of close 
cell types. While training some models, 
we observed that we can use the markers producing high sensitivity but at 
the same time can improve the specificity by increasing the probability 
threshold. Of course, this can only applied in some cases, because 
some markers can even have a larger affect on the specificity than the 
prediction probability threshold.*



# Save classification model for further use

New classification models can be stored using the `save_new_model` function:

```{r}
save_new_model(new_model = classifier_Stem,
               path_to_models = paste0(dir,"/scAnnotatR"),
               include.default = FALSE)

save_new_model(new_model = classifier_Intermediate,
               path_to_models = paste0(dir,"/scAnnotatR"),
               include.default = FALSE)

save_new_model(new_model = classifier_Colonocyte,
               path_to_models = paste0(dir,"/scAnnotatR"),
               include.default = FALSE)

save_new_model(new_model = classifier_BEST4,
               path_to_models = paste0(dir,"/scAnnotatR"),
               include.default = FALSE)

save_new_model(new_model = classifier_Goblet,
               path_to_models = paste0(dir,"/scAnnotatR"),
               include.default = FALSE)

save_new_model(new_model = classifier_Tuft,
               path_to_models = paste0(dir,"/scAnnotatR"),
               include.default = FALSE)
```


Parameters:

  * **new_model**: The new model that should be added to the database in the
                   specified directory.
  * **path_to_models**: The directory where the new models should be stored.
  * **include.default**: If set, the default models shipped with the package
                         are added to the database.

Users can also choose whether copy all pretrained models of the packages to the
new model database. If not, in the future, user can only choose to use either 
default pretrained models or new models by specifying only one path to models.

Models can be deleted from the model database using the `delete_model` function:

```{r}
# delete the "Intermediate cells" model from the new database
delete_model("Intermediate", path_to_models = paste0(dir,"/scAnnotatR"))
```
